---
title: "Running Nicheformer foundational model"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: >
    This package takes an h5ad file with the gene count matrix and spatial coordinates as input, runs Nicheformer, creating the Python environment automatically, and returns a matrix with the embeddings.
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
vignette: >
  %\VignetteIndexEntry{Running Nicheformer foundational model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

Loading the necessary packages to run `Nicheformer` in R

```{r setup}
library(anndataR)
library(fomo)
library(SpatialExperiment)
```

Preparing the annDataR object

```{r annDataR_creation}
spe <- readRDS(
  system.file("extdata", "CosMx1k_MouseBrain1_100tx_100cl.rds", package = "fomo")
)

spatialCoords(spe) |> as.data.frame() -> coords_df
colData(spe)$x_coord <- coords_df[, 1]
colData(spe)$y_coord <- coords_df[, 2]
reducedDim(spe, "spatial") <- as.matrix(spatialCoords(spe))

adata <- as_AnnData(spe)
colnames(adata$obsm$spatial) <- c("x_coord", "y_coord")
```

Here, we are using a mouse cosmx-based example with gene symbols as variable names. However, `Nicheformer` requires human Ensembl IDs. Therefore, this chunk retrieves and replaces the orthologues 

```{r orthologous}
library(homologene)
library(org.Hs.eg.db)
library(AnnotationDbi)

gene_symbols <- adata$var_names

# mouse (taxid 10090) -> human (taxid 9606) ortholog symbols
orthologs <- homologene(gene_symbols, inTax = 10090, outTax = 9606)
# columns: "10090" (mouse symbol), "9606" (human symbol)

# human symbol -> human Ensembl ID
human_ensembl <- mapIds(
  org.Hs.eg.db,
  keys      = orthologs[["9606"]],
  column    = "ENSEMBL",
  keytype   = "SYMBOL",
  multiVals = "first"
)

orthologs$ensembl_id <- human_ensembl[orthologs[["9606"]]]

# match back to original gene order
new_names <- ifelse(
  gene_symbols %in% orthologs[["10090"]],
  orthologs$ensembl_id[match(gene_symbols, orthologs[["10090"]])],
  gene_symbols
)

# keep only valid Ensembl IDs (start with ENSG) and not duplicated
is_ensembl  <- grepl("^ENSG", new_names)
is_unique   <- !duplicated(new_names)
keep        <- is_ensembl & is_unique

message(sum(keep), " genes kept out of ", length(new_names))

adata <- adata[, keep]$as_InMemoryAnnData()
rownames(adata$var) <- new_names[keep]
```

Generating the `h5ad` file and running `Nicheformer` in R

```{r running_nicheformer}
# write anndata to tempfile 
adata$write_h5ad(
  tp <- tempfile(fileext = ".h5ad"), 
  mode = "w"
)
nicheformer_data <- Run_nicheformer(adata_path = tp,
                                  technology = "cosmx")
```

Downstream analysis using the embeddings to calculate clusters using a graph-based approach

```{r clustering}
library(ggplot2)
library(uwot)
library(bluster)
library(cluster)     
library(patchwork)
library(dplyr)

set.seed(1234)

data_for_clustering <- nicheformer_data

leiden_clusters <- clusterRows(
    data_for_clustering,
    NNGraphParam(
        k             = 30,
        type          = "jaccard",        
        cluster.fun   = "leiden",
        cluster.args  = list(
            resolution_parameter = 1.0,  
            objective_function   = "modularity"
        )
    )
)

n_clusters <- length(unique(leiden_clusters))
message(sprintf("Leiden found %d clusters", n_clusters))

umap_2d <- umap(
    data_for_clustering,
    n_components = 2,
    n_neighbors  = 30,
    min_dist     = 0.3,
    metric       = "cosine"   
)

plot_df <- data.frame(
    umap1   = umap_2d[, 1],
    umap2   = umap_2d[, 2],
    x       = adata$obs$x_coord,
    y       = adata$obs$y_coord,
    cluster = factor(leiden_clusters)
)

p_umap <- ggplot(plot_df, aes(umap1, umap2, color = cluster)) +
    geom_point(size = 0.6, alpha = 0.7) +
    theme_classic() +
    labs(title = "Embedding space (UMAP)", x = "UMAP 1", y = "UMAP 2") +
    guides(color = guide_legend(override.aes = list(size = 3)))

p_spatial <- ggplot(plot_df, aes(x, y, color = cluster)) +
    geom_point(size = 0.6, alpha = 0.7) +
    coord_fixed() +
    theme_classic() +
    labs(title = "Physical space", x = "x", y = "y") +
    guides(color = guide_legend(override.aes = list(size = 3)))

p_umap + p_spatial + plot_layout(guides = "collect")
```

```{r sessionInfo}
sessionInfo()
```
