---
title: "Running Nicheformer foundational model"
author: ""
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: >
  This package takes an h5ad file with the gene count matrix and spatial coordinates as input, runs Nicheformer creating the Python environment automatically, and returns a matrix with the embeddings.
  
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
vignette: >
  %\VignetteIndexEntry{Running Nicheformer foundational model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

Loading the necessary packages to run `Nicheformer` in R

```{r setup}
library(anndataR)
library(fomo)
library(SpatialExperiment)
```

Preparing the AnnData object

```{r annDataR_creation}
spe <- readRDS(
  system.file("extdata", "CosMx1k_MouseBrain1_100tx_100cl.rds", package = "fomo")
)

adata <- as_AnnData(spe)

# add coordinates to metadata
spatialCoords(spe) |> as.data.frame() -> coords_df
adata$obs$x_coord <- coords_df[, 1]
adata$obs$y_coord <- coords_df[, 2]

# add coordinates as spatial
coords <- as.matrix(spatialCoords(spe))
rownames(coords) <- adata$obs_names
adata$obsm$spatial <- coords
colnames(adata$obsm$spatial) <- c("x_coord", "y_coord")
```

Here, we are using a mouse cosmx-based example with gene symbols as variable names. However, `Nicheformer` requires human Ensembl IDs. Therefore, this chunk retrieves and replaces the orthologues 

```{r orthologous}
library(homologene)
library(org.Hs.eg.db)
library(AnnotationDbi)

gene_symbols <- adata$var_names

# mouse (taxid 10090) -> human (taxid 9606) ortholog symbols
orthologs <- homologene(gene_symbols, inTax = 10090, outTax = 9606)
# columns: "10090" (mouse symbol), "9606" (human symbol)

# human symbol -> human Ensembl ID
human_ensembl <- mapIds(
  org.Hs.eg.db,
  keys      = orthologs[["9606"]],
  column    = "ENSEMBL",
  keytype   = "SYMBOL",
  multiVals = "first"
)

orthologs$ensembl_id <- human_ensembl[orthologs[["9606"]]]

# match back to original gene order
new_names <- ifelse(
  gene_symbols %in% orthologs[["10090"]],
  orthologs$ensembl_id[match(gene_symbols, orthologs[["10090"]])],
  gene_symbols
)

# keep only valid Ensembl IDs (start with ENSG) and not duplicated
is_ensembl  <- grepl("^ENSG", new_names)
is_unique   <- !duplicated(new_names)
keep        <- is_ensembl & is_unique

message(sum(keep), " genes kept out of ", length(new_names))

adata <- adata[, keep]$as_InMemoryAnnData()
rownames(adata$var) <- new_names[keep]
```

Generating the `h5ad` file and running `Nicheformer` in R

```{r running_nicheformer}
# write anndata to tempfile 
adata$write_h5ad(
  tp <- tempfile(fileext = ".h5ad"), 
  mode = "w"
)
nicheformer_data <- Run_nicheformer(adata_path = tp,
                                    technology = "cosmx")
reducedDim(spe, "nicheformer") <- nicheformer_data
```

Downstream analysis using the embeddings to calculate clusters using a graph-based approach

```{r clustering}
library(scrapper)

seed <- 1234
set.seed(seed)

g <- buildSnnGraph(t(nicheformer_data), 
                   num.neighbors = 30, 
                   weight.scheme = "jaccard")
leiden_clusters <- clusterGraph(g, 
                                method = "leiden", 
                                leiden.resolution = 1.0, 
                                leiden.objective = "modularity",
                                leiden.seed = seed)
spe$leiden_clusters <- leiden_clusters$membership

message(sprintf("Leiden found %d clusters", 
                length(unique(leiden_clusters))))

```

The resulting embedding of Nicheformer will still be of high dimensions, we can 
further reduce the dimensionality with PCA and UMAP. 

```{r embed_further}
# reduce with PCA
result_pca <- runPca(t(nicheformer_data), 
                     number = 30, 
                     scale = TRUE, 
                     seed = seed)
result_pca <- result_pca$components
reducedDim(spe, "nicheformer_PCA") <- t(result_pca)

# reduce with UMAP
result_umap <- runUmap(result_pca, 
                       optimize.seed = seed)
reducedDim(spe, "nicheformer_UMAP") <- result_umap
```

Let us visualize the UMAP and spatial observations

```{r vis}
library(ggspavis)
library(patchwork)
library(scater)

p_umap <- plotUMAP(spe, dimred = "nicheformer_UMAP")

spe$in_tissue <- 1
spe$x_centroid <- spe$y_centroid <- NULL
p_spatial <- plotCoords(spe, annotate = "leiden_clusters", point_size = 2)

p_umap + p_spatial
```

```{r sessionInfo}
sessionInfo()
```
