---
title: "Running scGPT with fomo"
package: fomo
output:
  BiocStyle::html_document
bibliography: ../inst/references.bib
vignette: >
  %\VignetteIndexEntry{Running scGPT with fomo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This vignette demonstrates how to use `Run_scGPT` to embed single-cell RNA-seq data using the scGPT model[@cui2024scgpt]. 

scGPT is a foundation model for single-cell mRNA-Seq data. scGPT can be used in many different context and strategies (see their official
[github](https://github.com/bowang-lab/scGPT) repository). In particular, scGPT provides multiple possible pretrained models which are linked to in their [github README](https://github.com/bowang-lab/scGPT/blob/main/README.md). The models differ based on the training data used. The most common choice of model is the whole-human [`scGPT_human`](https://drive.google.com/drive/folders/1oWh_-ZRdhtoGQ2Fw24HP41FgLoomVo-y), which is trained 33 million normal human cells across a range of tissue types. Tissue-type specific models are available. There is also a "continual pretrained model", [`scGPT_CP`](https://drive.google.com/drive/folders/1_GROJTzXiAV8HB4imruOTk6PEGuNOcgB) that  inherits the pre-trained scGPT whole-human model, and was further supervised by extra cell type labels (using the Tabula Sapiens dataset). 

In `Run_scGPT` we run a basic "zero-shot" embedding of the data via the `scg.tasks.embed_data()` function in the `scgpt` python package. "Zero-shot" means that the input data is run through the pretrained model, and the output embeddings are returned. Depending on which model is used, this could correspond to the [zero-shot tutorials of scGPT](https://github.com/bowang-lab/scGPT/tree/main/tutorials/zero-shot). Those tutorials also provide strategies for how to use these to perform integration of datasets. 

## Huggingface

scGPT also provides access to their model [huggingface model](https://huggingface.co/tdc/scGPT). We do not currently provide implementation of this model, since scGPT offers many more models than just that available on huggingface. We have focused on an implementation that allows users the ability to work with and compare these many different models.  However, this means users must download the models and store them in a local directory for `Run_scGPT`. 

# Prerequisites

`Run_scGPT` takes a `.h5ad` filename containing the cell transcripts data and a pretrained scGPT model directory as input, and returns a matrix of cell embeddings. We cache all files in a dedicated `model_dir` folder. 

```{r model_dir}
library(googledrive)
library(tools)
drive_deauth()

# create cache for fomo data
model_dir <- tools::R_user_dir("fomo/scgpt", which = "cache")
dir.create(model_dir, recursive = TRUE, showWarnings = FALSE)
```

## Input data

The input `.h5ad` file used in this vignette is the example data provided by scGPT (batch_covid_subsampled_train.h5ad )[https://github.com/bowang-lab/scGPT/blob/main/tutorials/zero-shot/Tutorial_ZeroShot_Reference_Mapping.ipynb], but we have subsetted to the first 100 cells so that it will run instantaneously even on a CPU, which we provide in a google drive and download the h5ad file from this location.

```{r h5ad}
url <- "https://drive.google.com/file/d/1Do7CXaaSTwEySGWHMKAkpN5G_g-OY8y2"
folder_id <- googledrive::as_id(url)
files <- drive_get(folder_id)
h5ad_file <- file.path(model_dir, files[1,]$name)
if(!file.exists(h5ad_file))
  drive_download(files[1,], path = h5ad_file)
```

## Model weights

The pretrained scGPT model directory that we use is `scGPT_human` model which can be downloaded from a google drive provided in https://github.com/bowang-lab/scGPT. The model directory should contain a `.pt` file with the model weights, along with `vocab.json` and `args.json`.

```{r model_weights}
url <- "https://drive.google.com/drive/folders/1oWh_-ZRdhtoGQ2Fw24HP41FgLoomVo-y"
folder_id <- googledrive::as_id(url)
files <- drive_ls(drive_get(folder_id), recursive = TRUE)
for (i in seq_along(files)){
  path <- file.path(model_dir, files[i,]$name)
  if(!file.exists(path))
    drive_download(files[i,], path = path)
}
list.files(model_dir)
```

# Running scGPT

Run `Run_scGPT` to embed the cells given the path to the h5ad file on disk `h5ad_file` and the mode directory `model_dir`. The result is a matrix with one row per cell and one column per embedding dimension:

```{r run}
library(fomo)
library(anndataR)
result <- Run_scGPT(
    h5ad_file = h5ad_file,
    model_dir  = model_dir,
    gene_col   = "gene_name"
)
dim(result)
```

The resulting embedding of scGPT will still be of high dimensions, we can 
further reduce the dimensionality with PCA and UMAP. 

```{r embed_further}
library(scrapper)

# reduce with PCA
result_pca <- runPca(t(result), number = 30, scale = TRUE)
result_pca <- result_pca$components

# reduce with UMAP
result_umap <- runUmap(result_pca)
```

Now we can add this new embeddings to a SingleCellExperiment generated from the 
same h5ad file. 

```{r embed}
library(SingleCellExperiment)
sce <- anndataR::read_h5ad(h5ad_file, as = "SingleCellExperiment")

reducedDim(sce, "scGPT") <- result
reducedDim(sce, "scGPT_PCA") <- t(result_pca)
reducedDim(sce, "scGPT_UMAP") <- result_umap

library(scater)
plotUMAP(sce, dimred = "scGPT_UMAP")
```

# Basic Python script

The basic python script that underlies this function is:

```
import scanpy as sc
import scgpt as scg
adata = sc.read_h5ad(h5ad_file)

ref_embed_adata = scg.tasks.embed_data(
    adata,
    model_dir,
    gene_col=gene_col,
    batch_size=64,
)
```

There is additional code added to ensure that it runs on MacOS. 

# Session Info

```{r sessionInfo}
sessionInfo()
```

# References

